---
title: "Getting Started with spectreasy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with spectreasy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

# spectreasy: Spectral Unmixing Quality Control

This vignette runs a small end-to-end `spectreasy` workflow with synthetic data.
The code is evaluated during vignette builds, so it demonstrates actual package
usage rather than static pseudocode.

```{r}
suppressPackageStartupMessages(library(spectreasy))
```

## Installation

Install the current development version with:

```r
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
remotes::install_github("pkheisig/spectreasy")

# Once accepted by Bioconductor:
# BiocManager::install("spectreasy")
```

## 1. Create a toy reference matrix

We start with a simple 3-marker reference matrix and derive a static OLS
unmixing matrix.

```{r}
M_demo <- rbind(
  FITC = c(1.00, 0.20, 0.05),
  PE   = c(0.10, 1.00, 0.20),
  APC  = c(0.05, 0.15, 1.00)
)
colnames(M_demo) <- c("B2-A", "YG1-A", "R1-A")

W_demo <- derive_unmixing_matrix(M_demo, method = "OLS")
W_demo
```

## 2. Build a toy `flowSet` in memory

A useful Bioconductor-friendly workflow is to keep multiple samples in memory as
a `flowSet`. The helper below simulates three single-stain-like samples.

```{r}
simulate_sample <- function(sample_name, dominant_marker, M, n_cells = 400) {
  markers <- rownames(M)
  marker_signal <- matrix(rexp(n_cells * length(markers), rate = 8), ncol = length(markers))
  colnames(marker_signal) <- markers
  marker_signal[, dominant_marker] <- rexp(n_cells, rate = 0.6) + 2

  raw_signal <- marker_signal %*% M + matrix(rnorm(n_cells * ncol(M), sd = 0.03), ncol = ncol(M))
  exprs_mat <- cbind(
    raw_signal,
    "Time" = seq_len(n_cells),
    "FSC-A" = rnorm(n_cells, mean = 90000, sd = 7000),
    "SSC-A" = rnorm(n_cells, mean = 45000, sd = 5000)
  )
  colnames(exprs_mat)[seq_len(ncol(M))] <- colnames(M)

  ff <- flowCore::flowFrame(exprs_mat)
  pd <- flowCore::pData(flowCore::parameters(ff))
  if (all(c("name", "desc") %in% colnames(pd))) {
    pd$desc <- pd$name
    flowCore::parameters(ff) <- methods::new("AnnotatedDataFrame", data = pd)
  }
  ff
}

toy_fs <- flowCore::flowSet(list(
  FITC_sample = simulate_sample("FITC_sample", "FITC", M_demo),
  PE_sample = simulate_sample("PE_sample", "PE", M_demo),
  APC_sample = simulate_sample("APC_sample", "APC", M_demo)
))

flowCore::sampleNames(toy_fs)
```

## 3. Unmix samples directly from the `flowSet`

`unmix_samples()` can work directly on an in-memory `flowSet`, which makes it
simpler to stay within the Bioconductor flow cytometry ecosystem.

```{r}
toy_unmixed <- unmix_samples(
  sample_dir = toy_fs,
  M = M_demo,
  method = "OLS",
  write_fcs = FALSE
)

names(toy_unmixed)
head(toy_unmixed[[1]]$data)
```

If you want an in-memory Bioconductor container back, set
`return_type = "flowSet"`.

```{r}
toy_unmixed_flowset <- unmix_samples(
  sample_dir = toy_fs,
  M = M_demo,
  method = "OLS",
  write_fcs = FALSE,
  return_type = "flowSet"
)

class(toy_unmixed_flowset)
flowCore::sampleNames(toy_unmixed_flowset)
```

`spectreasy` also supports `SingleCellExperiment`, which fits more naturally
into downstream Bioconductor single-cell workflows.

```{r}
if (requireNamespace("SingleCellExperiment", quietly = TRUE) &&
    requireNamespace("SummarizedExperiment", quietly = TRUE) &&
    requireNamespace("S4Vectors", quietly = TRUE)) {
  toy_exprs <- do.call(cbind, lapply(seq_along(toy_fs), function(i) {
    t(flowCore::exprs(toy_fs[[i]]))
  }))
  toy_coldata <- data.frame(
    sample_id = rep(flowCore::sampleNames(toy_fs), each = nrow(flowCore::exprs(toy_fs[[1]]))),
    row.names = paste0("cell_", seq_len(ncol(toy_exprs))),
    stringsAsFactors = FALSE
  )
  colnames(toy_exprs) <- rownames(toy_coldata)

  toy_sce <- suppressWarnings(
    SingleCellExperiment::SingleCellExperiment(
      assays = list(counts = toy_exprs),
      colData = S4Vectors::DataFrame(toy_coldata)
    )
  )

  toy_unmixed_sce <- suppressWarnings(
    unmix_samples(
      sample_dir = toy_sce,
      M = M_demo,
      method = "OLS",
      write_fcs = FALSE,
      return_type = "SingleCellExperiment"
    )
  )

  class(toy_unmixed_sce)
  SummarizedExperiment::assayNames(toy_unmixed_sce)
  SingleCellExperiment::altExpNames(toy_unmixed_sce)
}
```

## 4. Generate diagnostic plots without writing files

Most plotting helpers return `ggplot` objects and only save when you pass an
explicit `output_file`.

```{r}
plot_spectra(M_demo, output_file = NULL)
```

```{r}
plot_ssm(calculate_ssm(M_demo), output_file = NULL)
```

For SCC-like scatter diagnostics, provide a sample-to-marker mapping.

```{r}
sample_to_marker <- c(
  FITC_sample = "FITC",
  PE_sample = "PE",
  APC_sample = "APC"
)

plot_unmixing_scatter_matrix(
  unmixed_list = toy_unmixed,
  sample_to_marker = sample_to_marker,
  markers = rownames(M_demo),
  output_file = NULL,
  max_points_per_sample = 150,
  seed = 1
)
```

We can also summarize negative-population spread from the unmixed results.

```{r}
toy_results_df <- do.call(rbind, lapply(toy_unmixed, function(x) x$data))
plot_nps(calculate_nps(toy_results_df), output_file = NULL)
```

## 5. Build a QC report explicitly

Report functions write files only when you supply an explicit destination.
Here we generate a temporary PDF to show the full workflow.

```{r}
toy_report <- tempfile(fileext = ".pdf")
generate_qc_report(toy_results_df, M_demo, output_file = toy_report)
file.exists(toy_report)
```

## Real-data workflow

For real projects, organize your data like this:

```
my_project/
├── scc/                          # Single-color control FCS files
│   ├── FITC_beads.fcs
│   ├── PE_beads.fcs
│   └── Unstained.fcs             # Autofluorescence control
├── samples/                      # Experimental FCS files
│   └── Sample1.fcs
└── fcs_mapping.csv               # Control file (optional)
```

### Create or review the control file

```r
create_control_file(
  input_folder = "scc",
  cytometer = "Aurora",
  output_file = "fcs_mapping.csv"
)
```

### Generate the SCC review report

```r
generate_scc_report(
  scc_dir = "scc",
  control_file = "fcs_mapping.csv",
  cytometer = "Aurora",
  output_file = file.path("spectreasy_outputs", "SCC_QC_Report.pdf")
)
```

### Build the reference matrix

```r
control_df <- read.csv("fcs_mapping.csv", stringsAsFactors = FALSE, check.names = FALSE)

M <- build_reference_matrix(
  input_folder = "scc",
  output_folder = "gating_plots",
  save_qc_plots = TRUE,
  control_df = control_df,
  default_sample_type = "beads"
)
```

### Run control-stage unmixing

```r
ctrl <- autounmix_controls(
  scc_dir = "scc",
  control_file = "fcs_mapping.csv",
  cytometer = "Aurora",
  output_dir = "spectreasy_outputs/autounmix_controls"
)
```

### Launch the interactive GUI

```r
launch_gui(
  matrix_dir = "spectreasy_outputs/autounmix_controls",
  samples_dir = "samples"
)
```

### Unmix experimental samples

```r
unmixed <- unmix_samples(
  sample_dir = "samples",
  M = M,
  method = "WLS",
  output_dir = "spectreasy_outputs/unmix_samples",
  write_fcs = TRUE
)
```

## Function reference

| Function | Purpose |
|----------|---------|
| `generate_scc_report()` | Review SCC quality before unmixing |
| `build_reference_matrix()` | Extract signatures from single-color controls |
| `autounmix_controls()` | Build/save control-stage matrices and SCC unmix outputs |
| `unmix_samples()` | Unmix experimental FCS files or in-memory `flowSet`s |
| `launch_gui()` | Launch the interactive web interface |

```{r}
sessionInfo()
```
