---
title: "Getting Started with spectreasy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with spectreasy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  warning = FALSE,
  message = FALSE
)
```

```{r}
suppressPackageStartupMessages(library(spectreasy))
```

# Example workflow

This vignette demonstrates the primary `spectreasy` workflow on the packaged
example dataset. The example project contains:

- single-color controls in `scc/`
- one experimental sample in `sample/sample.fcs`

The user-facing workflow is:

1. download the example data into a project directory
2. run `autounmix_controls()`
3. review and supplement the generated `fcs_mapping.csv`
4. confirm the control file in the console so `autounmix_controls()` can finish
5. run `unmix_samples()`

## 1. Download the example data

`spectreasy_example_data()` downloads the example archive once, caches it under
the R user cache, and can copy the extracted files into a project directory for
local work.

```{r}
project_dir <- file.path(tempdir(), "spectreasy_vignette_project")
if (dir.exists(project_dir)) {
  unlink(project_dir, recursive = TRUE, force = TRUE)
}

example_paths <- spectreasy_example_data(dest_dir = project_dir)

list.files(project_dir, recursive = TRUE)
```

For the remainder of this vignette, the commands are shown as they would be run
from the project directory created above.

```{r, include = FALSE}
control_file <- file.path(project_dir, "fcs_mapping.csv")
auto_control_df <- create_control_file(
  input_folder = file.path(project_dir, "scc"),
  include_af_folder = FALSE,
  cytometer = "Aurora",
  output_file = file.path(project_dir, "fcs_mapping_auto.csv")
)

panel_annotations <- data.frame(
  filename = c(
    "Alexa Fluor 700 (Beads).fcs",
    "BUV395 (Beads).fcs",
    "BV510 (Beads).fcs",
    "FITC (Beads).fcs",
    "LIVE DEAD NIR (Cells).fcs",
    "PerCP-Cy5.5 (Beads).fcs",
    "Unstained (Cells).fcs"
  ),
  fluorophore = c(
    "Alexa Fluor 700",
    "BUV395",
    "BV510",
    "FITC",
    "LIVE DEAD NIR",
    "PerCP-Cy5.5",
    "AF"
  ),
  marker = c(
    "CD3",
    "CD45RA",
    "CD27",
    "CD8",
    "Live",
    "CCR7",
    "Autofluorescence"
  ),
  control.type = c("beads", "beads", "beads", "beads", "cells", "beads", "cells"),
  is.viability = c("", "", "", "", "TRUE", "", ""),
  stringsAsFactors = FALSE
)

match_idx <- match(auto_control_df$filename, panel_annotations$filename)
stopifnot(!anyNA(match_idx))

control_df <- auto_control_df
control_df$fluorophore <- panel_annotations$fluorophore[match_idx]
control_df$marker <- panel_annotations$marker[match_idx]
control_df$control.type <- panel_annotations$control.type[match_idx]
control_df$is.viability <- panel_annotations$is.viability[match_idx]

utils::write.csv(control_df, control_file, row.names = FALSE, quote = TRUE)

ctrl <- autounmix_controls(
  scc_dir = file.path(project_dir, "scc"),
  control_df = control_df,
  cytometer = "Aurora",
  output_dir = file.path(project_dir, "spectreasy_outputs", "autounmix_controls"),
  unmix_method = "WLS",
  build_qc_plots = TRUE,
  unmix_scatter_panel_size_mm = 30,
  seed = 1
)

unmixed <- unmix_samples(
  sample_dir = file.path(project_dir, "sample"),
  unmixing_matrix_file = file.path(
    project_dir,
    "spectreasy_outputs",
    "autounmix_controls",
    "scc_unmixing_matrix.csv"
  ),
  output_dir = file.path(project_dir, "spectreasy_outputs", "unmix_samples")
)
```

## 2. Start the control-stage workflow

Run `autounmix_controls()` first. If `fcs_mapping.csv` is missing,
`auto_create_control = TRUE` creates it automatically and then pauses for
review.

```{r, eval = FALSE}
setwd(project_dir)

autounmix_controls(
  scc_dir = "scc",
  auto_create_control = TRUE,
  cytometer = "Aurora",
  auto_unknown_fluor_policy = "by_channel",
  unmix_method = "WLS",
  build_qc_plots = TRUE,
  unmix_scatter_panel_size_mm = 30
)
```

After the control file is created, `autounmix_controls()` prints a confirmation
prompt and waits:

```text
Proceed with autounmix_controls now? [y/n]:
```

## 3. Review and supplement `fcs_mapping.csv`

Open the generated `fcs_mapping.csv` in the project directory and complete the
panel annotation before continuing. At minimum, review these columns:

- `fluorophore`
- `marker`
- `channel`
- `control.type`
- `is.viability`

For the example dataset used in this vignette, the reviewed control file looks
like this:

```{r, echo = FALSE}
knitr::kable(control_df)
```

## 4. Return to the console and confirm with `y`

Once `fcs_mapping.csv` has been reviewed and saved, return to the console where
`autounmix_controls()` is waiting and enter:

```text
y
```

The same `autounmix_controls()` call then continues and writes the control-stage
outputs to `spectreasy_outputs/autounmix_controls/`.

```{r, echo = FALSE}
list.files(file.path(project_dir, "spectreasy_outputs", "autounmix_controls"), recursive = TRUE)
```

Key outputs from this step include:

- `fcs_mapping.csv`
- `spectreasy_outputs/autounmix_controls/scc_reference_matrix.csv`
- `spectreasy_outputs/autounmix_controls/scc_unmixing_matrix.csv`
- `spectreasy_outputs/autounmix_controls/scc_unmixing_matrix.png`
- `spectreasy_outputs/autounmix_controls/scc_unmixing_scatter_matrix.png`
- `spectreasy_outputs/autounmix_controls/scc_unmixed/*.fcs`

## 5. Unmix the experimental sample

After the control-stage workflow has completed, unmix the experimental files
with `unmix_samples()`. The static unmixing matrix written by
`autounmix_controls()` is used directly.

```{r, eval = FALSE}
unmixed <- unmix_samples(
  sample_dir = "sample",
  unmixing_matrix_file = "spectreasy_outputs/autounmix_controls/scc_unmixing_matrix.csv",
  output_dir = "spectreasy_outputs/unmix_samples"
)
```

For the example dataset, this writes:

- `spectreasy_outputs/unmix_samples/sample_unmixed.fcs`

and returns a named list with one element per sample.

```{r, echo = FALSE}
sample_name <- names(unmixed)[1]
sample_results <- unmixed[[sample_name]]$data
preview_cols <- intersect(
  c("Alexa Fluor 700", "BUV395", "BV510", "FITC", "LIVE DEAD NIR", "PerCP-Cy5.5", "AF", "File"),
  colnames(sample_results)
)

knitr::kable(utils::head(sample_results[, preview_cols, drop = FALSE]))
```

```{r, echo = FALSE}
file.exists(file.path(project_dir, "spectreasy_outputs", "unmix_samples", "sample_unmixed.fcs"))
```

# Optional steps

The sections below are useful extensions, but they are not required for the
core `autounmix_controls()` -> `unmix_samples()` workflow shown above.

## Generate the SCC review report before unmixing

If you want a PDF review of the single-color controls before the control-stage
run, call `generate_scc_report()` first.

```{r}
scc_report_file <- file.path(project_dir, "spectreasy_outputs", "SCC_QC_Report.pdf")

generate_scc_report(
  scc_dir = file.path(project_dir, "scc"),
  control_df = control_df,
  cytometer = "Aurora",
  output_file = scc_report_file,
  seed = 1
)

file.exists(scc_report_file)
```

## Use a reviewed control table in non-interactive workflows

For scripts, reports, or CI jobs, you can pass a reviewed `control_df` directly
to `autounmix_controls()` and skip the confirmation prompt.

```{r}
ctrl_noninteractive <- autounmix_controls(
  scc_dir = file.path(project_dir, "scc"),
  control_df = control_df,
  cytometer = "Aurora",
  output_dir = file.path(project_dir, "spectreasy_outputs", "autounmix_controls_noninteractive"),
  unmix_method = "WLS",
  build_qc_plots = FALSE,
  seed = 1
)

dim(ctrl_noninteractive$M)
```

## Unmix directly from the reference matrix

Instead of using the saved static matrix file, you can unmix dynamically from
the in-memory reference matrix returned by `autounmix_controls()`.

```{r}
fluor_reference_matrix <- ctrl$M
marker_map <- stats::setNames(control_df$marker, control_df$fluorophore)
reference_matrix <- fluor_reference_matrix
rownames(reference_matrix) <- unname(marker_map[rownames(reference_matrix)])

unmixed_dynamic <- unmix_samples(
  sample_dir = file.path(project_dir, "sample"),
  M = reference_matrix,
  method = "WLS",
  output_dir = file.path(project_dir, "spectreasy_outputs", "unmix_samples_dynamic")
)

names(unmixed_dynamic)
```

## Inspect quick QC plots

Reference spectra and spectral spread plots should be interpreted in
fluorophore space, so the original fluorophore-labeled control matrix is used
here.

```{r}
reference_matrix_no_af <- fluor_reference_matrix[!grepl("^AF($|_)", rownames(fluor_reference_matrix), ignore.case = TRUE), , drop = FALSE]

plot_spectra(reference_matrix_no_af, output_file = NULL)
```

```{r}
plot_ssm(calculate_ssm(reference_matrix_no_af), output_file = NULL)
```

```{r}
plot_nps(calculate_nps(sample_results), output_file = NULL)
```
