---
title: "Getting Started with spectreasy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with spectreasy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  warning = FALSE,
  message = FALSE
)
```

# spectreasy: spectral unmixing on example FCS files

This vignette runs a complete `spectreasy` workflow on example FCS files stored
as a GitHub release asset. `spectreasy_example_data()` downloads the archive
matching the installed package version on first use, unzips it automatically,
and returns local paths to:

- single-color controls in `scc/`
- one experimental sample in `sample/sample.fcs`

All code is evaluated during vignette builds.

```{r}
suppressPackageStartupMessages(library(spectreasy))
```

## Installation

For the released Bioconductor version, install with:

```r
BiocManager::install("spectreasy")
```

If you need the current development version from GitHub, install with:

```r
remotes::install_github("pkheisig/spectreasy")
```

`remotes` is used here rather than `devtools` because only GitHub installation is needed.

## 1. Download the example data and prepare a temporary project

We first download the example-data archive from the GitHub release matching the
installed package version. `spectreasy_example_data()` caches the zip file and
unzips it automatically.
We then copy the extracted `sample/` and `scc/` folders into a temporary
working directory so that the vignette writes outputs outside the cache.

```{r}
example_paths <- spectreasy_example_data(force=TRUE)

project_dir <- file.path(tempdir(), "spectreasy_vignette_project")
if (dir.exists(project_dir)) {
  unlink(project_dir, recursive = TRUE, force = TRUE)
}
dir.create(project_dir, recursive = TRUE)
dir.create(file.path(project_dir, "scc"))
dir.create(file.path(project_dir, "sample"))

file.copy(
  list.files(example_paths$scc_dir, full.names = TRUE, pattern = "\\.[Ff][Cc][Ss]$"),
  file.path(project_dir, "scc")
)
file.copy(
  list.files(example_paths$sample_dir, full.names = TRUE, pattern = "\\.[Ff][Cc][Ss]$"),
  file.path(project_dir, "sample")
)

list.files(project_dir, recursive = TRUE)
```

## 2. Create and review `fcs_mapping.csv`

We first let `spectreasy` infer the detector channels from the example controls,
then we replace the fluorophore and marker annotations with the intended panel
labels for this vignette.

```{r}
control_file <- file.path(project_dir, "fcs_mapping.csv")

auto_control_df <- create_control_file(
  input_folder = file.path(project_dir, "scc"),
  include_af_folder = FALSE,
  cytometer = "Aurora",
  output_file = file.path(project_dir, "fcs_mapping_auto.csv")
)

manual_annotations <- data.frame(
  filename = c(
    "Alexa Fluor 700 (Beads).fcs",
    "BUV395 (Beads).fcs",
    "BV510 (Beads).fcs",
    "FITC (Beads).fcs",
    "LIVE DEAD NIR (Cells).fcs",
    "PerCP-Cy5.5 (Beads).fcs",
    "Unstained (Cells).fcs"
  ),
  fluorophore = c(
    "Alexa Fluor 700",
    "BUV395",
    "BV510",
    "FITC",
    "LIVE DEAD NIR",
    "PerCP-Cy5.5",
    "AF"
  ),
  marker = c(
    "CD3",
    "CD45RA",
    "CD27",
    "CD8",
    "Live",
    "CCR7",
    "Autofluorescence"
  ),
  control.type = c("beads", "beads", "beads", "beads", "cells", "beads", "cells"),
  is.viability = c("", "", "", "", "TRUE", "", ""),
  stringsAsFactors = FALSE
)

match_idx <- match(auto_control_df$filename, manual_annotations$filename)
stopifnot(!anyNA(match_idx))

control_df <- auto_control_df
control_df$fluorophore <- manual_annotations$fluorophore[match_idx]
control_df$marker <- manual_annotations$marker[match_idx]
control_df$control.type <- manual_annotations$control.type[match_idx]
control_df$is.viability <- manual_annotations$is.viability[match_idx]

utils::write.csv(control_df, control_file, row.names = FALSE, quote = TRUE)
knitr::kable(control_df)
```

## 3. Review the single-color controls

Before unmixing, we generate the SCC review report.

```{r}
scc_report_file <- file.path(project_dir, "spectreasy_outputs", "SCC_QC_Report.pdf")

generate_scc_report(
  scc_dir = file.path(project_dir, "scc"),
  control_df = control_df,
  cytometer = "Aurora",
  output_file = scc_report_file,
  seed = 1
)

file.exists(scc_report_file)
```

## 4. Build the control-stage outputs

`autounmix_controls()` builds the reference matrix, saves control-stage plots,
and unmixes the controls.

```{r}
ctrl <- autounmix_controls(
  scc_dir = file.path(project_dir, "scc"),
  control_df = control_df,
  cytometer = "Aurora",
  output_dir = file.path(project_dir, "spectreasy_outputs", "autounmix_controls"),
  build_qc_plots = FALSE,
  seed = 1
)

dim(ctrl$M)
rownames(ctrl$M)
```

The control-stage matrix is labeled by fluorophore, which is useful for control
QC. For downstream sample interpretation, we relabel the matrix rows to marker
names using `fcs_mapping.csv`.

```{r}
marker_map <- stats::setNames(control_df$marker, control_df$fluorophore)
reference_matrix <- ctrl$M
rownames(reference_matrix) <- unname(marker_map[rownames(reference_matrix)])

rownames(reference_matrix)
```

We can visualize the non-autofluorescence spectra directly from the evaluated
matrix.

```{r}
reference_matrix_no_af <- reference_matrix[rownames(reference_matrix) != "Autofluorescence", , drop = FALSE]
plot_spectra(reference_matrix_no_af, output_file = NULL)
```

```{r}
plot_ssm(calculate_ssm(reference_matrix_no_af), output_file = NULL)
```

## 5. Unmix the example sample

Now we unmix `sample/sample.fcs` using the marker-labeled reference matrix.

```{r}
unmixed <- unmix_samples(
  sample_dir = file.path(project_dir, "sample"),
  M = reference_matrix,
  method = "WLS",
  output_dir = file.path(project_dir, "spectreasy_outputs", "unmix_samples"),
  write_fcs = TRUE
)

sample_results <- unmixed$sample$data
knitr::kable(utils::head(sample_results[, c(
  "CD3", "CD45RA", "CD27", "CD8", "CCR7", "Live", "Autofluorescence", "File"
)]))
```

The unmixed FCS file is written only because we explicitly supplied an output
folder and set `write_fcs = TRUE`.

```{r}
file.exists(file.path(project_dir, "spectreasy_outputs", "unmix_samples", "sample_unmixed.fcs"))
```

## 6. Summarize sample-level diagnostics

We can inspect the negative-population spread on the evaluated sample output.

```{r}
plot_nps(calculate_nps(sample_results), output_file = NULL)
```

And we can generate the sample QC PDF report.

```{r}
qc_report_file <- file.path(project_dir, "spectreasy_outputs", "QC_Report.pdf")
generate_qc_report(sample_results, reference_matrix, output_file = qc_report_file)
file.exists(qc_report_file)
```

## 7. Generated files

The evaluated vignette produces the expected control-stage and sample-stage
artifacts.

```{r}
output_files <- list.files(file.path(project_dir, "spectreasy_outputs"), recursive = TRUE)
knitr::kable(data.frame(file = output_files, stringsAsFactors = FALSE))
```

The interactive GUI can be launched later from the generated control outputs,
but it is not started during vignette builds.

```{r}
sessionInfo()
```
