---
title: "Getting Started with spectreasy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with spectreasy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  warning = FALSE,
  message = FALSE
)
```

```{r}
suppressPackageStartupMessages(library(spectreasy))
```

# Example workflow

This vignette demonstrates the primary `spectreasy` workflow on the packaged
example dataset. The example project contains:

- single-color controls in `scc/`
- one experimental sample in `sample/sample.fcs`

The user-facing workflow is:

1. download the example data into a project directory
2. run `unmix_controls()`
3. review and supplement the generated `fcs_mapping.csv`
4. confirm the control file in the console so `unmix_controls()` can finish
5. run `unmix_samples()`
6. generate QC reports for the controls and unmixed samples (optional)

## 1. Download the example data

`spectreasy_example_data()` downloads the example archive once, caches it under
the R user cache, and can copy the extracted files into a project directory for
local work.

```{r}
project_dir <- file.path(tempdir(), "spectreasy_vignette_project")
if (dir.exists(project_dir)) {
  unlink(project_dir, recursive = TRUE, force = TRUE)
}

example_paths <- spectreasy_example_data(dest_dir = project_dir)

list.files(project_dir, recursive = TRUE)
```

For the remainder of this vignette, the commands are shown as they would be run
from the project directory created above.

```{r, include = FALSE}
control_file <- file.path(project_dir, "fcs_mapping.csv")
auto_control_df <- create_control_file(
  input_folder = file.path(project_dir, "scc"),
  include_af_folder = FALSE,
  cytometer = "Aurora",
  output_file = file.path(project_dir, "fcs_mapping_auto.csv")
)

panel_annotations <- data.frame(
  filename = c(
    "Alexa Fluor 700 (Beads).fcs",
    "BUV395 (Beads).fcs",
    "BV510 (Beads).fcs",
    "FITC (Beads).fcs",
    "LIVE DEAD NIR (Cells).fcs",
    "PerCP-Cy5.5 (Beads).fcs",
    "Unstained (Cells).fcs"
  ),
  fluorophore = c(
    "Alexa Fluor 700",
    "BUV395",
    "BV510",
    "FITC",
    "LIVE DEAD NIR",
    "PerCP-Cy5.5",
    "AF"
  ),
  marker = c(
    "CD3",
    "CD45RA",
    "CD27",
    "CD8",
    "Live",
    "CCR7",
    "Autofluorescence"
  ),
  control.type = c("beads", "beads", "beads", "beads", "cells", "beads", "cells"),
  is.viability = c("", "", "", "", "TRUE", "", ""),
  stringsAsFactors = FALSE
)

match_idx <- match(auto_control_df$filename, panel_annotations$filename)
stopifnot(!anyNA(match_idx))

control_df <- auto_control_df
control_df$fluorophore <- panel_annotations$fluorophore[match_idx]
control_df$marker <- panel_annotations$marker[match_idx]
control_df$control.type <- panel_annotations$control.type[match_idx]
control_df$is.viability <- panel_annotations$is.viability[match_idx]

utils::write.csv(control_df, control_file, row.names = FALSE, quote = TRUE)

ctrl <- unmix_controls(
  scc_dir = file.path(project_dir, "scc"),
  control_file = control_file,
  cytometer = "Aurora",
  output_dir = file.path(project_dir, "spectreasy_outputs", "unmix_controls"),
  unmix_method = "WLS",
  unmix_scatter_panel_size_mm = 30,
  seed = 1
)

unmixed <- unmix_samples(
  sample_dir = file.path(project_dir, "sample"),
  unmixing_matrix_file = file.path(
    project_dir,
    "spectreasy_outputs",
    "unmix_controls",
    "scc_reference_matrix.csv"
  ),
  output_dir = file.path(project_dir, "spectreasy_outputs", "unmix_samples")
)
```

## 2. Start the control-stage workflow

Run `unmix_controls()` first. If `fcs_mapping.csv` is missing,
`auto_create_control = TRUE` creates it automatically and then pauses for
review.

```{r, eval = FALSE}
setwd(project_dir)

unmix_controls(
  scc_dir = "scc",
  auto_create_control = TRUE,
  cytometer = "Aurora",
  auto_unknown_fluor_policy = "by_channel",
  unmix_method = "WLS",
  unmix_scatter_panel_size_mm = 30
)
```

After the control file is created, `unmix_controls()` prints a confirmation
prompt and waits:

```text
Proceed with unmix_controls now? [y/n]:
```

## 3. Review and supplement `fcs_mapping.csv`

Open the generated `fcs_mapping.csv` in the project directory and complete the
panel annotation before continuing. At minimum, review these columns:

- `fluorophore`
- `marker`
- `channel`
- `control.type`
- `is.viability`

For the example dataset used in this vignette, the reviewed control file looks
like this:

```{r, echo = FALSE}
knitr::kable(control_df)
```

## 4. Return to the console and confirm with `y`

Once `fcs_mapping.csv` has been reviewed and saved, return to the console where
`unmix_controls()` is waiting and enter:

```text
y
```

The same `unmix_controls()` call then continues and writes the control-stage
outputs to `spectreasy_outputs/unmix_controls/`.

```{r, echo = FALSE}
list.files(file.path(project_dir, "spectreasy_outputs", "unmix_controls"), recursive = TRUE)
```

Key outputs from this step include:

- `fcs_mapping.csv`
- `spectreasy_outputs/unmix_controls/scc_reference_matrix.csv`
- `spectreasy_outputs/unmix_controls/scc_variances.csv`
- `spectreasy_outputs/unmix_controls/scc_unmixing_matrix.csv`
- `spectreasy_outputs/unmix_controls/scc_unmixing_matrix.png`
- `spectreasy_outputs/unmix_controls/scc_unmixing_scatter_matrix.png`
- `spectreasy_outputs/unmix_controls/fsc_ssc/*.png`
- `spectreasy_outputs/unmix_controls/histogram/*.png`
- `spectreasy_outputs/unmix_controls/spectrum/*.png`
- `spectreasy_outputs/unmix_controls/scc_unmixed/*.fcs`

The control-stage run also writes visual checks for each single-color control.
For one color, the three plots below show the FSC/SSC gate, the peak-channel
histogram gate, and the detector spectrum used to build the reference matrix.

```{r, echo = FALSE, out.width = "48%", fig.show = "hold"}
example_control <- "FITC (Beads)"
control_plot_dir <- file.path(project_dir, "spectreasy_outputs", "unmix_controls")
knitr::include_graphics(c(
  file.path(control_plot_dir, "fsc_ssc", paste0(example_control, "_fsc_ssc.png")),
  file.path(control_plot_dir, "histogram", paste0(example_control, "_histogram.png"))
))
```

```{r, echo = FALSE, out.width = "100%"}
knitr::include_graphics(
  file.path(control_plot_dir, "spectrum", paste0(example_control, "_spectrum.png"))
)
```

The same run creates the NxN scatter matrix for the single-color controls. Each
row is one control, and each column checks how much signal appears in the other
unmixed markers.

```{r, echo = FALSE, out.width = "100%"}
knitr::include_graphics(ctrl$unmixing_scatter_file)
```

### What WLS uses

When `method = "WLS"`, `spectreasy` uses detector noise measured from the
single-color controls. It does not guess those weights from the reference matrix
or from sample brightness.

That means WLS needs two pieces:

- the reference spectra in `scc_reference_matrix.csv`
- the SCC-derived noise estimates in `scc_variances.csv`

`unmix_controls()` writes both files. If `unmix_samples(method = "WLS")` can
load the reference matrix but cannot find `scc_variances.csv`, it tries to
recompute the missing variances from the SCC files. This is still true WLS,
because the weights still come from the controls. If neither a variance file nor
the SCC files are available, WLS stops instead of silently falling back to a
guess.

The `control.type` column in `fcs_mapping.csv` also matters here. It tells
`spectreasy` whether each control should be gated as `beads` or `cells`. If
`control.type` is empty, `spectreasy` falls back to filename-based guessing.

## 5. Unmix the experimental sample

After the control-stage workflow has completed, unmix the experimental files
with `unmix_samples()`. The reference matrix written by
`unmix_controls()` is loaded by default.

```{r, eval = FALSE}
unmixed <- unmix_samples(
  sample_dir = "sample",
  output_dir = "spectreasy_outputs/unmix_samples"
)
```

For the example dataset, this writes:

- `spectreasy_outputs/unmix_samples/sample_unmixed.fcs`

and returns a named list with one element per sample.

```{r, echo = FALSE}
sample_name <- names(unmixed)[1]
sample_results <- unmixed[[sample_name]]$data
preview_cols <- intersect(
  c("Alexa Fluor 700", "BUV395", "BV510", "FITC", "LIVE DEAD NIR", "PerCP-Cy5.5", "AF", "File"),
  colnames(sample_results)
)

knitr::kable(utils::head(sample_results[, preview_cols, drop = FALSE]))
```

```{r, echo = FALSE}
file.exists(file.path(project_dir, "spectreasy_outputs", "unmix_samples", "sample_unmixed.fcs"))
```

## 6. Generate quality control reports (optional)

After unmixing, you can generate comprehensive PDF reports to inspect the quality of both the single-color controls and the unmixed experimental samples.

### Single-Color Control (SCC) Report

The SCC report reviews gating, peak channels, and signal distributions for each control file.

```{r}
scc_report_file <- file.path(project_dir, "spectreasy_outputs", "SCC_QC_Report.pdf")

generate_scc_report(
  scc_dir = file.path(project_dir, "scc"),
  cytometer = "Aurora",
  output_file = scc_report_file,
  seed = 1
)

file.exists(scc_report_file)
```

### Samples Report

The overall sample report visualizes unmixing quality across samples, including spectra overlays, detector residuals, spread matrices, and marker scatter plots.

```{r}
qc_report_file <- file.path(project_dir, "spectreasy_outputs", "Sample_QC_Report.pdf")

generate_sample_report(
  results = unmixed,
  M = ctrl$M,
  output_file = qc_report_file
)

file.exists(qc_report_file)
```

# Optional steps

The sections below are useful extensions, but they are not required for the
core `unmix_controls()` -> `unmix_samples()` workflow shown above.

## Per-cell Autofluorescence (AF) Extraction

By default, `unmix_controls()` and `unmix_samples()` use one average
autofluorescence signature from the unstained control. If your cells have
different AF shapes from cell to cell, you can split AF into several basis
signatures.

Use the two multi-AF settings in the control-stage call. `af_n_bands` controls
how many AF basis signatures are extracted from the unstained control, while
`include_multi_af` tells `spectreasy` to include additional AF controls from
the `af/` directory when those files are available.

```{r, eval = FALSE}
ctrl_multi_af <- unmix_controls(
  scc_dir = "scc",
  control_file = "fcs_mapping.csv",
  cytometer = "Aurora",
  output_dir = "spectreasy_outputs/unmix_controls_multi_af",
  unmix_method = "WLS",
  include_multi_af = TRUE,
  af_n_bands = 10,
  seed = 1
)
```

If you want `unmix_samples()` to rebuild the reference matrix dynamically from
the SCC files, pass the same two AF settings there as well:

```{r, eval = FALSE}
unmixed_multi_af <- unmix_samples(
  sample_dir = "sample",
  unmixing_matrix_file = NULL,
  scc_dir = "scc",
  control_file = "fcs_mapping.csv",
  method = "WLS",
  include_multi_af = TRUE,
  af_n_bands = 10,
  output_dir = "spectreasy_outputs/unmix_samples_multi_af"
)
```

`af_n_bands` is like choosing how many AF "flavors" to model. More bands can
fit complex AF better, but they also take longer to compute. For large files,
start with 5 to 10 bands before trying higher values.

## Use a reviewed control CSV in non-interactive workflows

For scripts, reports, or CI jobs, you can supply a pre-existing, reviewed control CSV file via `control_file` to `unmix_controls()` to skip the confirmation prompt.

```{r}
ctrl_noninteractive <- unmix_controls(
  scc_dir = file.path(project_dir, "scc"),
  control_file = control_file,
  cytometer = "Aurora",
  output_dir = file.path(project_dir, "spectreasy_outputs", "unmix_controls_noninteractive"),
  unmix_method = "WLS",
  seed = 1
)

dim(ctrl_noninteractive$M)
```

## Pass the in-memory reference matrix directly

You can pass the in-memory reference matrix returned by `unmix_controls()`
directly to `unmix_samples()` instead of loading it from the saved CSV file.
For WLS, this in-memory matrix already carries the SCC-derived variances with
it. If you save only the matrix CSV and reload it later, keep the matching
`scc_variances.csv` next to it or pass it through `variances_file`.

```{r}
fluor_reference_matrix <- ctrl$M
marker_map <- stats::setNames(control_df$marker, control_df$fluorophore)
reference_matrix <- fluor_reference_matrix
mapped_names <- marker_map[rownames(reference_matrix)]
na_idx <- is.na(mapped_names)
mapped_names[na_idx] <- rownames(reference_matrix)[na_idx]
rownames(reference_matrix) <- unname(mapped_names)

unmixed_dynamic <- unmix_samples(
  sample_dir = file.path(project_dir, "sample"),
  M = reference_matrix,
  method = "WLS",
  output_dir = file.path(project_dir, "spectreasy_outputs", "unmix_samples_dynamic")
)

names(unmixed_dynamic)
```

## Inspect quick QC plots

Reference spectra and spectral spread plots should be interpreted in
fluorophore space, so the original fluorophore-labeled control matrix is used
here.

```{r}
reference_matrix_no_af <- fluor_reference_matrix[!grepl("^AF($|_)", rownames(fluor_reference_matrix), ignore.case = TRUE), , drop = FALSE]

plot_spectra(reference_matrix_no_af, output_file = NULL)
```

```{r}
plot_ssm(calculate_ssm(reference_matrix_no_af), output_file = NULL)
```

```{r}
plot_nps(calculate_nps(sample_results), output_file = NULL)
```

## Session information

```{r}
sessionInfo()
```
